---
title: "MovieLens"
author: "Daniel Teodoro"
date: "9/22/2019"
output: pdf_document
---

```{r setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)
if(!require(tidyverse)) install.packages("tidyverse", repos = "http://cran.us.r-project.org")
if(!require(caret)) install.packages("caret", repos = "http://cran.us.r-project.org")
```

## Project 2 - MovieLens
## Daniel Teodoro

# Generates predicted movie ratings and calculates RMSE.

```{r prepare, echo=FALSE}
vPath <- paste(getwd(), "/", sep="")
setwd(vPath)
vMovieLensFile <- paste(vPath, "ml-10M100K/movielens.csv", sep="")
vCached <- file.exists(vMovieLensFile)
```


```{r acquire_data, echo=FALSE}
if (!vCached)
{
  
  # Zip
  vZIPFile <- paste(vPath, "ml-10m.zip", sep="")
  download.file("http://files.grouplens.org/datasets/movielens/ml-10m.zip", vZIPFile)
  unlink(paste(vPath, "ml-10M100K", sep=""), recursive = TRUE)
  
  # Ratings
  vRatingsFile <- "ml-10M100K/ratings.dat"
  unzip(vZIPFile, vRatingsFile)
  vRatingsDS <- readLines(vRatingsFile)
  ratings <- read.table(text = gsub("::", "\t", vRatingsDS), col.names = c("userId", "movieId", "rating", "timestamp"))
  rm(vRatingsDS)
  
  # Movies
  vMoviesFile <- "ml-10M100K/movies.dat"
  unzip(vZIPFile, vMoviesFile)
  vMoviesDS <- readLines(vMoviesFile)
  movies <- str_split_fixed(vMoviesDS, "\\::", 3)
  rm(vMoviesDS)
  colnames(movies) <- c("movieId", "title", "genres")
  movies <- as.data.frame(movies) %>% mutate(movieId = as.numeric(levels(movieId))[movieId],
                                             title = as.character(title),
                                             genres = as.character(genres))
  #MovieLens
  movielens <- left_join(ratings, movies, by = "movieId")
  
  # Cache results
  write.csv(movielens, vMovieLensFile)  
  
  rm(vZIPFile, vRatingsFile, vMoviesFile)
} else {  
  movielens <- read.csv(vMovieLensFile, row.names = 1)
}
rm(vCached, vPath, vMovieLensFile)
```



```{r process, echo=FALSE}

# Validation set will be 1/4 of MovieLens data
set.seed(1)
test_index <- createDataPartition(y = movielens$rating, times = 1, p = 1/4, list = FALSE)
edx <- movielens[-test_index,]  # Training Set
temp <- movielens[test_index,]  # Validation Set

# Make sure userId and movieId in validation set are also in edx set
validation <- temp %>% 
  semi_join(edx, by = "movieId") %>%
  semi_join(edx, by = "userId")

# Add rows removed from validation set back into edx set
removed <- anti_join(temp, validation)
edx <- rbind(edx, removed)

# Run algorithm on validation set to generate ratings
validation <- validation %>% select(-rating)

```


## Quiz

# Q1) How many rows and columns are there in the edx dataset?

```{r quiz_q1, echo=FALSE}
paste('The edx dataset has',nrow(edx),'rows and',ncol(edx),'columns.')
```

# Q2) How many zeros and threes were given in the edx dataset?

```{r quiz_q2, echo=FALSE}
paste(sum(edx$rating == 0), 'ratings with 0 were given and',
      sum(edx$rating == 3),'ratings with 3')
```

# Q3) How many different movies are in the edx dataset?

```{r quiz_q3, echo=FALSE}
edx %>% summarize(n_movies = n_distinct(movieId))
```

# Q4) How many different users are in the edx dataset?

```{r quiz_q4, echo=FALSE}
edx %>% summarize(n_users = n_distinct(userId))
```

# Q5) How many movie ratings are in each of the following genres in the edx dataset?

```{r quiz_q5, echo=FALSE}
drama <- edx %>% filter(str_detect(genres,"Drama"))
paste('Drama has',nrow(drama),'movies')

comedy <- edx %>% filter(str_detect(genres,"Comedy"))
paste('Comedy has',nrow(comedy),'movies')

thriller <- edx %>% filter(str_detect(genres,"Thriller"))
paste('Thriller has',nrow(thriller),'movies')

romance <- edx %>% filter(str_detect(genres,"Romance"))
paste('Romance has',nrow(romance),'movies')

rm(drama, comedy, thriller, romance)
```

# Q6) Which movie has the greatest number of ratings?

```{r quiz_q6, echo=FALSE}
edx %>% group_by(title) %>% summarise(number = n()) %>% arrange(desc(number))
```

# Q7) What are the five most given ratings in order from most to least?

```{r quiz_q7, echo=FALSE}
head(sort(-table(edx$rating)),5)
```

# Q8) True or False: 
In general, half star ratings are less common than whole star ratings 
(e.g., there are fewer ratings of 3.5 than there are ratings of 3 or 4, etc.).

```{r quiz_q8, echo=FALSE}
ratings35 <- table(edx$rating)["3.5"]
ratings3 <- table(edx$rating)["3"]
ratings4 <- table(edx$rating)["4"]
answer <- (ratings35 < ratings3 && ratings35 < ratings4)
print(answer)
rm(ratings35, ratings3, ratings4, answer)
```


## Data Analysis

# More than 10 million ratings.

```{r analysis_1, echo=FALSE}
str(movielens)
```

# The plot shows that 5 ratings  are more common that 0.5.

```{r analysis_2, echo=FALSE}
hist(movielens$rating)
summary(movielens$rating)
```

# Recent movies get more ratings. 

```{r analysis_3, echo=FALSE}
movielens$year <- as.numeric(substr(as.character(movielens$title),nchar(as.character(movielens$title))-4,nchar(as.character(movielens$title))-1))
plot(table(movielens$year))
```


# Last decades, ratings average get lower.

```{r analysis_4, echo=FALSE}
avg_ratings <- movielens %>% group_by(year) %>% summarise(avg_rating = mean(rating))
plot(avg_ratings)
```

## Results

# RMSE

$\sqrt{\overline{(TrueRatings - PredictedRatings) ^ 2}}$


```{r results, echo=FALSE}

# Root Mean Square Error
RMSE <- function(true_ratings, predicted_ratings)
  {
    sqrt(mean((true_ratings - predicted_ratings)^2))
  }

adj_factors <- seq(0, 5, 0.5)

rmses <- sapply(adj_factors, function(l){
  
  # The mean of training set
  mts <- mean(edx$rating)
  
  # Grade down low number on ratings
  me <- edx %>% 
    group_by(movieId) %>%
    summarize(me = sum(rating - mts)/(n()+l))

  # Derive penalty value 'adj_factor'
  predicted_ratings <- 
    edx %>% 
    left_join(me, by = "movieId") %>%
    mutate(pred = mts + me) %>%
    .$pred
  
  return(RMSE(predicted_ratings, edx$rating))
})

plot(adj_factors, rmses)

adj_factor <- adj_factors[which.min(rmses)]
paste('Best RMSE:',min(rmses),'is achieved with Adjustment Factor:',adj_factor)
```

## Predictions

# CSV file <predictions.csv>

```{r prediction}

predictions <- sapply(adj_factor,function(l){
  
  # The mean of training set
  mts <- mean(edx$rating)
  
  # Get movie effect with best adjustment factor
  me <- edx %>% 
    group_by(movieId) %>%
    summarize(me = sum(rating - mts)/(n()+l))
  
  # Predict on validation set
  predicted_ratings <- 
    validation %>% 
    left_join(me, by = "movieId") %>%
    mutate(pred = mts + me) %>%
    .$pred 
  
  return(predicted_ratings)
  
})

write.csv(validation %>% select(userId, movieId) %>% mutate(rating = predictions), 
          "predictions.csv", na = "", row.names=FALSE)

```


## Conclusion

# Using 1/4 as part of the data for the MovieLens validation set and removing the low rating 
# records in the training model, it was possible to obtain an RMSE index of 0.94 and, 
# verifying it with actual data, the model was reasonably efficient.


# References
http://www.montana.edu/rotella/documents/502/MarkdownEqnExamples.Rmd
https://www.calvin.edu/~rpruim/courses/s341/S17/from-class/MathinRmd.html
http://www.datainsight.at/report/Capstone_04_fin.html
https://r4ds.had.co.nz
http://stackoverflow/
